biological data preperation

### 2018-09-19 src/biological_data_prep.Rmd
NAO data added

>>>>

#NAO values

The data has been prepared as per the environmental_data_preperation.Rmd

```{r}
nao <- read.csv("../output/env/nao.csv", header=TRUE)
head(nao)
```

change month col names to just numeric
```{r}
names(nao)[2:13] = c("1", "2","3", "4", "5", "6", "7", "8", "9", "10", "11", "12")
head(nao)
```

Start with the month data first...subset cols 1:13 (where the year and month data are held)

```{r}
nao_yrmt <- nao[ ,1:13]
nao_yrmt
```

```{r}
nao_yrmt <- reshape(nao_yrmt, 
  varying = c("1", "2","3", "4", "5", "6", "7", "8", "9", "10", "11", "12"),
  v.names = "value",
  direction = "long")

nao_yrmt <- nao_yrmt[order(amo_yrmt$year),]
nao_yrmt
```

rename time to month.. and value to nao_sample

```{r}
colnames(nao_yrmt)[2] <- "month"
colnames(nao_yrmt)[3] <- "nao_sample"
nao_yrmt
```

now match the year and the month in the two dataframes, and populate the species dataset (data_aea) with nao_sample

```{r}
data_aea <- merge(x = data_aea, y = nao_yrmt[ , c("year", "month", "nao_sample")], by = c("year", "month"), all.x = TRUE)
data_aea
```

ok next - NAO value at previous sampling month/year

shift the nao_sample column down by 1...

```{r}
library(useful)
nao_yrmt <- shift.column(data = nao_yrmt, columns = "nao_sample", up = FALSE)
nao_yrmt
```

Note the shifted values are in a new column called amo_sample.Shifted. rename to amo_prev

```{r}
colnames(nao_yrmt)[colnames(nao_yrmt)=="nao_sample.Shifted"] <- "nao_prev"
nao_yrmt
write.csv(nao_yrmt, "../output/env/nao_prev.csv", row.names = FALSE) #for some reason i need to write to a csv and reload for the next step to work properly
```

now grab the data...

```{r}
nao_prev <- read.csv("../output/env/nao_prev.csv")
data_aea <- merge(x = data_aea, y = nao_prev[ , c("year", "month", "nao_prev")], by = c("year", "month"), all.x = TRUE)
data_aea
```

and now - NAO phase at previous winter (because the winter values are thought to be a major driver of ocean conditions in the following spring, summer, and autumn)

You need to extract the seasonal values from the NAO dataset

```{r}
nao_winter <- subset(nao, select = c("year", "WinterAvg"))
nao_winter
```

now shift the winter average data

```{r}
nao_winter <- shift.column(data = nao_winter, columns = "WinterAvg", up = FALSE)
nao_winter
write.csv(nao_winter, "../output/env/nao_winter.csv", row.names = FALSE) #for some reason need to save to csv then reload in the next step...

```

remember your shifted column is called WinterAvg.Shifted

now extract the WinterAvg.Shifted values to the species dataset

```{r}
nao_winter <- read.csv("../output/env/nao_winter.csv")
data_aea <- merge(x = data_aea, y = nao_winter[ , c("year", "WinterAvg.Shifted")], by = c("year"), all.x = TRUE)
colnames(data_aea)[colnames(data_aea)=="WinterAvg.Shifted"] <- "nao_winter" #change the colname
data_aea
```

ok and write a csv

```{r}
write.csv(data_aea,"../output/bio/data_aea_cell_amo_nao.csv", row.names = FALSE)
```
